Working title -- A heterogeneous landscape of selection and epistasis in protein-coding genes

### 2022-02-22 Manuscript.Rmd
add numerical checks

---
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: no
    number_sections: false
    keep_tex: true
    includes:
      in_header: "preamble.tex"
    citation_package: natbib
    extra_dependencies: ["float"]
bibliography: "references.bib"
biblio-style: "genetics"
title: Can we distinguish modes of selective interactions using linkage disequilibrium?
author:
header-includes:
- |
  ```{=latex}
  \usepackage{authblk}
  \author{Aaron P. Ragsdale \\ aaronpeaceragsdale@gmail.com}
  \affil{Department of Human Genetics, McGill University, Montreal, Canada \\
         National Laboratory of Genomics for Biodiversity, Irapuato, Mexico \\
         }
  ```
date: \today
---

@@

Selected mutations interfere and interact with evolutionary processes at nearby
loci, distorting allele frequency trajectories and correlations between pairs
of mutations. A number of recent studies have used patterns of linkage
disequilibrium (LD) between selected variants to test for selective
interference and epistatic interactions, with some disagreement over
interpreting observations from data. Interpretation is hindered by the relative
lack of analytic or even numerical expectations for patterns of variation
between pairs of loci under the combined effects of selection, dominance,
epistasis, and demography. Here, I develop a numerical approach to compute the
expected two-locus sampling distribution under diploid selection with arbitrary
epistasis and dominance, recombination, and variable population size. I use
this to explore how epistasis and dominance affect expected signed LD,
including for non-steady-state demography relevant to human populations.
Finally, I use whole-genome sequencing data from humans to assess how well we
can differentiate modes of selective interactions in practice. I find that
positive LD between missense mutations within genes is driven by strong
positive allele-frequency correlations between pairs of mutations that fall
within the same conserved domain, pointing to compensatory mutations or
antagonistic epistasis as the prevailing mode of interaction within but not
outside of conserved genic elements. The heterogeneous landscape of both
mutational fitness effects and selective interactions within protein-coding
genes calls for more refined inferences of the joint distribution of fitness
and interactive effects, and the methods presented here should prove useful in
that pursuit.

@@

Most new mutations that affect fitness are deleterious and tend to be
eliminated from a population. The amount of time that a deleterious mutation
segregates depends on the strength of selection against genomes that carry it,
with very damaging mutations kept at low frequencies and purged relatively
rapidly. In the time between mutation and fixation or loss, selected variants,
both beneficial and damaging, can dramatically impact patterns of variation in
nearby linked regions. This distortion away from neutral expectations is now
empirically documented using sequencing data from an ever-growing set of study
systems, but questions remain about the primary mode of interactions between
multiple linked variants and their joint effects on genome-wide patterns of
diversity.

@@

In this paper, I develop a numerical approach to solve for the two-locus
sampling distribution under a general diploid selection model with variable
recombination and single-population size history. I use this model to describe
how epistasis and dominance shape expected patterns of signed LD, under both
steady-state and non-equilibrium demography, that have been used to test for
interference and epistasis in population genomic data. I then turn to human
sequencing data and compare patterns of LD for synonymous, missense, and
loss-of-function mutations within and between protein-coding genes and
annotated domains. I show that while synonymous and missense variants display
similar slightly positive average LD within genes, for missense mutations this
signal is driven by correlations between pairs of mutations within, but not
between, protein-coding domains, suggesting an importance for antagonistic
epistasis or a prevalence of compensatory nonsynonymous mutations within
conserved elements. Finally, I discuss how well we can expect to discriminate
between modes of selective interactions using genome-wide patterns of LD.

@@

The most direct way to test for interactions between linked selected variants
is through deep mutation scanning experiments, in which many distinct mutations
are introduced within a target gene and then organismal fitness or some protein
function is experimentally measured [@Romero2009-yi; @Bank2015-vq;
@Puchta2016-rx; @Steinberg2016-is]. For example, using the model system of the
TEM-1 $\beta$-lactamase gene in *E. coli*, @Bershtein2006-bk found evidence for
synergistic epistasis, where multiple deleterious mutations had a greater
effect on fitness than the multiplication of the observed effects of the
individual mutations. The scale of mutation scanning experiments continues to
improve dramatically, promising greater resolution of the fitness landscape in
such model systems that can be compared to evolutionary theory
[@Otwinowski2018-zk].

@@

In the Methods, I expand on the moment system developed in @Ragsdale2019-nt to
compute the expected sampling distribution $\Psi_n$ of two-locus haplotypes
under a general model of selective interactions. This sampling distribution
stores the expected density or observed counts of pairs of biallelic loci with
each possible haplotype frequency configuration in a sample of size $n$. In
many cases it is simpler to visualize summaries such as the expectation or
variance of $D$ or slices of the distribution (e.g., Figures \@ref(fig:fig1),
\@ref(fig:fig2) and \@ref(fig:hudson)) instead of the full three-dimensional
distribution. Here I focus on low-order LD statistics including $\E[D]$ and
$\E[D^2]$ and their decay with recombination distance, as these are statistics
that are commonly used to test for interactions between loci. Instead of
unnormalized $\E[D^2]$ and $\E[D]$ I consider expectations for $\sigma_d^2 =
\E[D^2]/\E[p(1-p)q(1-q)]$ and $\sigma_d^1 = \E[D]/\E[p(1-p)q(1-q)]$, where $p$
and $q$ are the derived allele frequencies at the left ($A/a$) and right
($B/b$) loci, respectively. These normalized statistics have two benefits:
first, the mutation rate cancels so that expectations are robust to assumptions
about the per-base mutation rate, and second, analytic expectations for these
quantities are available under neutrality at constant population size for
comparison [@Ohta1971-yd].

@@

(ref:fig1cap) \textbf{Additive selection and epistasis.} Left panels show
expectations for the decay of $\sigma_d^2 = \E[D^2] / \E[p(1-p)q(1-q)]$ with
recombination distance, and right panels show expectations for $\sigma_d^1 =
\E[D] / \E[p(1-p)q(1-q)]$. The scaled recombination rate is given by
$\rho=4N_er$, where $r$ is the probability of recombination between two loci.
All expectations are at stationarity, and dashed lines are expectations under
neutrality (Equation \@ref(eq:ohta)). (A-B) Additive selection, with equal
selection strength at both loci. Stronger selection reduces $\sigma_d^2$ (A),
and in (B), we see that the negative LD effect of Hill-Robertson interference
is strongest for intermediate selection ($\gamma \approx -1$ to $-5$, where
$\gamma=4N_es$). (C-F) Epistasis with additive selection. Synergistic epistasis
(\(\epsilon < 0\)) decreases $\sigma_d^2$ while antagonistic epistasis
(\(\epsilon > 0\)) increases it compared to expectations under a multiplicative
fitness model ($\epsilon=0$). (See Methods for the definition of the
epistasis model.) Epistasis also pushes signed LD ($\sigma_d^1$) away from
zero, with synergistic epistasis decreasing LD for negatively selected
mutations and antagonistic epistasis increasing LD. The deviation of
$\sigma_d^1$ from zero under epistasis can extend for long recombination
distances, particularly for stronger selection.

@@

With epistasis, mean signed LD is large for both weakly and strongly selected
variants, with sign depending on the direction of epistatic interactions
(Figure \@ref(fig:fig1)). Synergistic epistasis results in negative LD while
antagonistic epistasis results in positive LD, and large positive LD can occur
even when the epistatic effect is relatively weak. Epistasis-induced LD can
extend over long distances, especially for strongly deleterious mutations. For
example, in Figure \@ref(fig:fig1)F even moderately deleterious mutations with
population-size-scaled selection coefficients of $\gamma=-10$ show large mean
LD that extends well beyond 100 kilobases. More strongly deleterious
interacting mutations are expected to show large signed LD over much larger
recombination distances.

@@

The effects of epistasis on LD have been widely studied. However, the combined
effect of non-additive selection within a locus and interference has received
less attention. @Garcia2020-zn used large-scale forward simulations to explore
the effect of dominance on patterns of LD, showing that LD depends on both the
magnitude of the selection and dominance coefficients in a nonlinear way, but
note that expensive large-scale simulations are needed to explore these effects
as no closed form approach to computing those expectations exist. The method
presented here provides such a closed form approach, so that expectations of
signed LD and other two-locus statistics can be computed more efficiently and
accurately.

@@

(ref:fig2cap) **Dominance effects.** (A and C) Pairs of recessive mutations
($h<0.5$) have higher squared LD than additive mutations, while pairs of
dominant mutations have lower expected $D^2$. (B and D) Pairs of recessive
deleterious mutations can have large negative LD, although this correlation
decays more rapidly with larger recombination than LD under epistasis (Figure
\@ref(fig:fig1)D and F). Notably, there is no simple relationship between
strength of selection and dominance, particularly when $s$ is $O(1/N)$. Note
that this setup does not include epistatic effects (i.e., $\epsilon=0$).

@@

Analytic progress beyond these results has come haltingly. In the
1980s, recursions were developed to compute the two-locus sampling distribution
under neutrality, that is, the probability of observing given counts of
two-locus haplotypes in a sample of size $n$ [@Golding1984-pu]. This approach
would continue to be developed and later form the foundation for the inference
of local recombination rates from population genetic data [@Hudson2001-sg;
@McVean2004-gj].

>>>>

---
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: no
    number_sections: false
    keep_tex: true
    includes:
      in_header: "preamble.tex"
    citation_package: natbib
    extra_dependencies: ["float"]
bibliography: "references.bib"
biblio-style: "genetics"
title: Working title -- A heterogeneous landscape of selection and epistasis in protein-coding genes
author:
header-includes:
- |
  ```{=latex}
  \usepackage{authblk}
  \author{Aaron P. Ragsdale \\ apragsdale@wisc.edu}
  \affil{Department of Integrative Biology, University of Wisconsin--Madison, WI \\
         Department of Human Genetics, McGill University, Montreal, Canada \\
         National Laboratory of Genomics for Biodiversity, Irapuato, Mexico \\
         }
  ```
- \usepackage{lineno}
- \linenumbers
date: \today
---

@@

Selected mutations interfere and interact with evolutionary processes at nearby
loci, distorting allele frequency trajectories and creating correlations between pairs
of mutations. A number of recent studies have used patterns of linkage
disequilibrium (LD) between selected variants to test for selective
interference and epistatic interactions, with some disagreement over
interpreting observations from data. Interpretation is hindered by the relative
lack of analytic or even numerical expectations for patterns of variation
between pairs of loci under the combined effects of selection, dominance,
epistasis, and demography. Here, I develop a numerical approach to compute the
expected two-locus sampling distribution under diploid selection with arbitrary
epistasis and dominance, recombination, and variable population size. I use
this to explore how epistasis and dominance affect expected signed LD,
including for non-steady-state demography relevant to human populations.
Finally, I use whole-genome sequencing data from humans to assess how well we
can differentiate modes of selective interactions in practice. I find that
positive LD between missense mutations within genes is driven by strong
positive allele-frequency correlations between pairs of mutations that fall
within the same conserved domain, pointing to compensatory mutations or
antagonistic epistasis as the prevailing mode of interaction within but not
outside of conserved genic elements. The heterogeneous landscape of both
mutational fitness effects and selective interactions within protein-coding
genes calls for more refined inferences of the joint distribution of fitness
and interactive effects, and the methods presented here should prove useful in
that pursuit.

@@

Most new mutations that affect fitness are deleterious and tend to be
eliminated from a population **Citations**. The amount of time that a deleterious mutation
segregates depends on the strength of selection against genomes that carry it,
with very damaging mutations kept at low frequencies and purged relatively
rapidly. In the time between mutation and fixation or loss, selected variants,
both beneficial and damaging, can dramatically impact patterns of variation in
nearby linked regions **Citations**.
This distortion away from neutral expectations is now
empirically documented using sequencing data from an ever-growing set of study
systems, but questions remain about the primary mode of interactions between
multiple linked variants and their joint effects on genome-wide patterns of
diversity.

@@

In this paper, I develop a numerical approach to solve for the two-locus
sampling distribution under a general diploid selection model with variable
recombination and single-population size history. I use this model to describe
how epistasis and dominance shape expected patterns of signed LD, under both
steady-state and non-equilibrium demography, that have been used to test for
interference and epistasis in population genomic data. I then turn to human
sequencing data and compare patterns of LD for synonymous, missense, and
loss-of-function mutations within and between protein-coding genes and
annotated domains. I show that while synonymous and missense variants display
similar slightly positive average LD within genes, for missense mutations this
signal is driven by correlations between pairs of mutations within, but not
between, protein-coding domains, suggesting an importance for antagonistic
(or "diminishing-returns")
epistasis or a prevalence of compensatory nonsynonymous mutations within
conserved elements. Finally, I discuss how well we can expect to discriminate
between modes of selective interactions using genome-wide patterns of LD.

@@

The most direct way to test for interactions between linked selected variants
is through deep mutation scanning experiments, in which many distinct mutations
are introduced within a target gene and then organismal fitness or some protein
function is experimentally measured [@Romero2009-yi; @Bank2015-vq;
@Puchta2016-rx; @Steinberg2016-is].
For example, using the model system of the TEM-1 $\beta$-lactamase gene in
*E. coli*, @Bershtein2006-bk found evidence for synergistic epistasis,
in which the combined effect of multiple deleterious mutations 
on individual fitness was larger than would be expected from multiplying
the independent observed effects of each individual mutation.
The scale of mutation scanning experiments continues to
improve dramatically, promising greater resolution of the fitness landscape in
such model systems that can be compared to evolutionary theory
[@Otwinowski2018-zk].

@@

In the Methods, I expand on the moment system developed in @Ragsdale2019-nt to
compute the expected sampling distribution $\Psi_n$ of two-locus haplotypes
under a general model of selective interactions. This sampling distribution
stores the expected density or observed counts of pairs of biallelic loci with
each possible haplotype frequency configuration in a sample of size $n$.
Below, we compute expectations for $\Psi_n$ under varying scenarios of
selection and interaction between pairs of loci.
It is not possible in this framework to include the effects of additional
selection mutations, such as background selection due to many linked variants.
Individual-based forward simulations are still needed for such scenarios
(e.g., Figure \@ref(fig:fwdpy11)).

In many cases it is simpler to visualize summaries such as the expectation or
variance of $D$ or slices of the distribution (e.g., Figures \@ref(fig:fig1),
\@ref(fig:fig2) and \@ref(fig:hudson)) instead of the full three-dimensional
distribution. Here I focus on low-order LD statistics including $\E[D]$ and
$\E[D^2]$ and their decay with recombination distance, as these are statistics
that are commonly used to test for interactions between loci.
For pairs of biallelic loci, with alleles $A/a$ at the left locus and $B/b$ at the
right locus, $D=f_{AB}-f_A f_B$ is the standard covariance measure of LD,
where $f_{AB}$ is the frequency of haplotypes carrying both $A$ and $B$, and
$f_A$ and $f_B$ are the marginal allele frequencies of $A$ and $B$.

Instead of unnormalized $\E[D^2]$ and $\E[D]$ I consider expectations for
$\sigma_d^2 = \E[D^2]/\E[f_A(1-f_A)f_B(1-f_B)]$ and
$\sigma_d^1 = \E[D]/\E[f_A(1-f_A)f_B(1-f_B)]$.
Such normalized statistics have two benefits:
first, the mutation rate cancels so that expectations are robust to assumptions
about the per-base mutation rate, and second, analytic expectations for these
quantities are available under neutrality at constant population size for
comparison [@Ohta1971-yd].

@@

(ref:fig1cap) \textbf{Additive selection and epistasis.}
Left panels show
expectations for the decay of $\sigma_d^2 = \E[D^2] / \E[p(1-p)q(1-q)]$ with
recombination distance, and right panels show expectations for $\sigma_d^1 =
\E[D] / \E[p(1-p)q(1-q)]$. The scaled recombination rate is given by
$\rho=4N_er$, where $r$ is the probability of recombination between two loci.
All expectations are at stationarity, and dashed lines are expectations under
neutrality (Equation \@ref(eq:ohta)). (A-B) Additive selection, with equal
selection strength at both loci. Stronger selection reduces $\sigma_d^2$ (A),
and in (B), we see that the negative LD effect of Hill-Robertson interference
is strongest for intermediate selection ($\gamma \approx -1$ to $-5$, where
$\gamma=4N_es$). (C-F) Epistasis with additive selection. Synergistic epistasis
(\(\epsilon > 0\)) decreases $\sigma_d^2$ while antagonistic epistasis
(\(\epsilon < 0\)) increases $\sigma_d^2$
compared to expectations under a multiplicative
fitness model ($\epsilon=0$). (See Methods for the definition of the
epistasis model.) Epistasis also pushes signed LD ($\sigma_d^1$) away from
zero, with synergistic epistasis decreasing LD for negatively selected
mutations and antagonistic epistasis increasing LD. The deviation of
$\sigma_d^1$ from zero under epistasis can extend for long recombination
distances, particularly for stronger selection.

@@

With epistasis, mean signed LD is large for both weakly and strongly selected
variants, with sign depending on the direction of epistatic interactions
(Figure \@ref(fig:fig1)).
Synergistic epistasis (in which the effect of two mutations together is larger than
the product of each individual mutation's effect) results in negative LD while
antagonistic epistasis (in which the combined effect is less than the product of
independent effects) results in positive LD, and large positive LD can occur
even when the epistatic effect is relatively weak. Epistasis-induced LD can
extend over long distances, especially for strongly deleterious mutations. For
example, in Figure \@ref(fig:fig1)F even moderately deleterious mutations with
population-size-scaled selection coefficients of $\gamma=-10$ show large mean
LD that extends to values of $\rho$ much greater than 1 (for humans, assuming
roughly 1 cM/Mb, this is on the order of 100 kilobases or more).
More strongly deleterious interacting mutations are expected to show large
signed LD over much larger recombination distances.

@@

The effects of epistasis on LD have been widely studied. However, the combined
effect of non-additive selection within a locus and interference has received
less attention. @Garcia2020-zn used large-scale forward simulations to explore
the effect of dominance on patterns of LD, showing that LD depends on both the
magnitude of the selection and dominance coefficients in a nonlinear way, but they
note that expensive large-scale simulations are needed to explore these effects
as no closed form approach to computing those expectations exist.
The numerical method presented here provides such a closed form approach, so that expectations of signed LD and other two-locus statistics can be computed more
efficiently and accurately.

@@

(ref:fig2cap) **The effects of dominance on LD.**
(A and C) Pairs of recessive mutations
($h<0.5$) have higher expected squared LD than additive mutations, while pairs of
dominant mutations have lower expected $\sigma_d^2$.
(B and D) Pairs of recessive
deleterious mutations can have large negative LD, although this correlation
decays more rapidly with larger recombination than LD under epistasis (Figure
\@ref(fig:fig1)D and F). Notably, there is no simple relationship between
strength of selection and dominance, particularly when $s$ is $O(1/N)$. Note
that the expected values shown here do not include epistatic effects
(i.e., $\epsilon=0$).

@@

Analytic progress beyond these results has come haltingly. In the 1980s,
recursions were developed to compute the two-locus sampling distribution under
neutrality, that is, the probability of observing given counts of two-locus
haplotypes in a sample of size $n$ [@Golding1984-pu]. This approach would
continue to be developed and later form the foundation for the inference of
local recombination rates from population genetic data [@Hudson2001-sg;
@McVean2004-gj]. More recently, @Song2007-qk computed $\E[r^2]$ using a
diffusion approximation approach, although their solution involves the
summation of infinitely many terms and is restricted to neutrality and
steady-state demography.
